A Speculative Approach to Parallelization in Particle Swarm Optimization

### 2011-04-01 masters_thesis/si_paper/paper.tex
Done?

In this paper we show that the results of standard PSO can be reproduced
\emph{exactly}, two iterations at a time, using a speculative approach adapted
from speculative execution. We show that the standard PSO equations can be
factored such that a set of speculative positions can be found which will
always include the position computed in the next iteration.  By computing the
value of the objective function for each of the speculative positions at the
same time the algorithm evaluates the objective function for the current
position, it is possible to know the objective function values for both the
current and the next iteration at the same time.  We demonstrate this principle
by implementation and show that it produces exactly the same results as
standard PSO, but two iterations at a time.  The resulting implementation runs
efficiently on large clusters where the number of processors is much larger
than a typical or reasonable number of particles, producing better results in
less wall-clock time.

@@

\section{Implementing Speculative Evaluation}
\label{sec:implementation}

It is not trivial in some parallel architectures to determine which speculative
position was the correct next position of each particle.  In this section we
move from the mathematics of our method to its implementation.  First we
discuss the relatively easy case of a centralized parallel PSO algorithm with a
master computer and many slaves.  In such an architecture, the master keeps
track of all necessary information with only trivial message passing needed.
Then we discuss the more complicated case of a distributed algorithm, where
each particle is on its own and needs to send and receive messages to and from
other particles.  Finally we discuss the further complications of a dynamic
topology such as Random, where a particle's neighbors change from one iteration
to another.

In distributed frameworks, synchronizing all of the machines for a round of
message passing can be expensive.  The method just described uses three rounds
of message passing for every two iterations (corresponding to steps 2, 5 and 10
in \alg{centralized}).  It is possible to perform speculative evaluation in PSO
with only one round of communication per two iterations.  However, the
methodology is tedious and is not the focus of this paper, so we defer its
description to the Appendix.

@@

While the idea of speculative evaluation in particle swarm optimization is
intriguing, the results in \sec{results} show that in many cases it requires
too many extra evaluations to produce competitive results.  On most functions,
the larger swarm size that can be used in standard parallelizations leads to
better performance than is found by trying to do speculative evaluation to
exactly reproduce PSO.  However, if we keep the idea of speculative evaluation
while relaxing the requirement of exactly reproducing the behavior of the
original PSO algorithm, we see some impressive results.

@@

To make better use of the extra speculative evaluations, instead of choosing
the speculative child that matches the branch that the original PSO would have
taken, we take the child that has the best value.  The methodology is exactly
the same as above except for the process of choosing which speculative child to
accept.  The only change needed in \alg{centralized} is in step 7, where
$\noeval{\s}_{t+1}$ with the best value is chosen from $\noeval{\sset}_{t+1}$
instead of with the matching branch.  We call this technique Pick Best.

@@

Because the problem facing speculative evaluation is having too many possible
evaluations to perform, a natural step to take is to eliminate some of the
evaluations.  If we could reliably predict which branch were going to be taken,
we could limit ourselves to one speculative evaluation per particle instead of
$2n+1$.  With a fixed number of processors, this would allow us to greatly
increase the swarm size relative to that needed in the original speculative
algorithm (e.g., with 120 processors, a standard parallelization has a swarm of
size 120, complete speculative evaluation has a swarm of size 15, and pruning
the evaluations to only one per particle allows a swarm of size 60).  As not
all of the branches are evaluated in any given iteration, we call this
technique pruning.

@@

In parallel PSO on long-running functions, the natural way to present results
is in terms of iterations, not function evaluations.  This is because when all
function evaluations at each iteration are performed concurrently, iterations
are the direct equivalent of wall-clock time.  It is true that with a fixed
number of processors, function evaluations are also equivalent to wall-clock
time.  However, reporting function evaluations in a paper on parallel PSO can
be misleading if compared to a paper on serial PSO, as the actual wall-clock
times corresponding to the function evaluations are not directly comparable.
Thus we report iterations in our results instead of function evaluations.  But
because SEPSO actually performs two iterations of PSO at each ``iteration,''
and Social Promotion and Many Iterations make the idea of ``iterations of PSO''
somewhat nebulus, we instead call each ``iteration'' a ``time step.''

@@

When performing pruning in Pick Best Pruned and Social Promotion Pruned, there
are a large number of ways to prune speculative evaluations.  We experimented
with several, but present results for only one possible pruning.  The pruning
we present uses only the two branches where the $\nbest$ value was not updated:
\casepn\ and \casexn.  Those branches are convenient in that no messages are
needed from neighbors in order to produce the positions of the speculative
particles---in distributed frameworks using several rounds of communication
(see \sec{distributed}), one of the rounds of communication can be dropped
entirely.  Pruning all but these branches also allows the use of arbitrarily
dense topologies, as the number of speculative particles is no longer dependent
on the number of neighbors the particle has.

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{sphere20}
  \caption{Function Sphere with a swarm that uses 240 processors per time step,
  comparing a pertinent subset of possible methods.}
  \label{fig:sphere-20}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{rastrigin}
  \caption{Function Rastrigin with a swarm that uses 240 processors per time
  step.  Recall that at each time step all processors evaluate the function
  once.  All speculative algorithms use the Random topology.}
  \label{fig:rastrigin}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank6}
  \caption{Function Griewank with a swarm that uses 800 processors per time
  step.  Recall that at each time step all processors evaluate the function
  once.  Instead of showing average function value, we show the percentage of
  runs that have found the global optimum by each iteration.  All algorithms
  use the Ring topology.}
  \label{fig:basic-griewank1}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank1}
  \caption{Function Griewank with a swarm that uses 240 processors per time
  step.  All algorithms use the Ring topology.}
  \label{fig:basic-griewank2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank3}
  \caption{Function Griewank with a swarm that uses 240 processors per time
  step, comparing Pick Best with results from \fig{basic-griewank2}.  All
  algorithms use the Ring topology.}
  \label{fig:griewank-pickbest}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank4}
  \caption{Function Griewank with a swarm that uses 240 processors per time
  step, comparing pruning with the best results from \fig{griewank-pickbest}.
  All algorithms use the Ring topology.}
  \label{fig:griewank-pruned}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank5}
  \caption{Function Griewank with a swarm that uses 240 processors per time
  step, comparing speculating several iterations ahead with selections from
  \fig{griewank-pruned}.  All algorithms use the Ring topology.}
  \label{fig:griewank-manyiters}
\end{figure}

@@

Bohachevsky is a unimodal function best optimized with a Complete swarm.  It is
similar to Griewank in that there is a global optimum with a value of 0, and
the swarm sometimes finds it and sometimes does not.  Thus we present a graph
similar to those of Griewank, because ``average value'' graphs have a
misleading flat line.  In \fig{bohachevsky} we show a plot with a few pertinent
methods, while \tabref{bohachevsky-20} shows all of the results.  All of our
speculative approaches found the optimum much quicker than Standard with a
Random topology.  However, SEPSO was slower than Standard Complete and got
stuck 25\% of the time.  Pick Best, Pick Best Pruned, and Many Iterations all
outperformed Standard Complete, with Many Iterations finding the optimum about
twice as fast.

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{bohachevsky}
  \caption{Function Bohachevsky with a swarm that uses 480 processors per time
  step.  Recall that at each time step all processors evaluate the function
  once.}
  \label{fig:bohachevsky}
\end{figure}

@@

\begin{table}
  \caption{Average speed up, comparing the number of time steps to completion
  for the best speculative topology to the time steps to completion for the
  best standard topology.  All functions have 50 dimensions.}
  \label{tab:50dims-summary}
  \centering
  \begin{tabular}{|l|r|}
	\hline
	Function&Speed Up Factor\\
	\hline
	Rastrigin&N/A\\
	\hline
	Sphere&1.84\\
	\hline
	Schwefel&2.51\\
	\hline
	Griewank&6.08\\
	\hline
	Bohachevsky&3.07\\
	\hline
  \end{tabular}
\end{table}

@@

At 500 dimensions we found that the performance of constricted PSO on benchmark
functions becomes rather dismal.  In almost all cases, Standard fails to make
any significant progress.  Our search for an explanation seems to say that the
space is too large for particles to converge to the same point, so all updates
to the best position found come from a single particle wandering in the space.
Every time that single particle finds a new best position, its velocity
contracts, so the particle is unable to make significant progress on its own.
In constricted PSO it is necessary to have a collection of particles exploring
a promising location to keep the particles' velocities from contracting
prematurely, and in 500 dimensions the space is too large to get the collection
of particles to the same location.

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{sphere500}
  \caption{Results for the 500 dimensional Sphere function.  We use 800
  processors for this experiment, so each time step corresponds to 800 function
  evaluations.}
  \label{fig:sphere500}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank500}
  \caption{Results for the 500 dimensional Griewank function.  We use 800
  processors for this experiment, so each time step corresponds to 800 function
  evaluations.}
  \label{fig:griewank500}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{rastrigin500}
  \caption{Results for the 500 dimensional Rastrigin function.  We use 800
  processors for this experiment, so each time step corresponds to 800 function
  evaluations.}
  \label{fig:rastrigin500}
\end{figure}

@@

\section*{Appendix---Alternate form of message passing}

>>>>

In this paper we show that the results of standard PSO can be reproduced
\emph{exactly}\footnote{In fact it is only because the results are exactly the
same that we are confident of our implementation.  With the careful use of
random seeds we were able to detect errors in our speculative implementation
when particle positions were off in the tenth digit.}, two iterations at a
time, using a speculative approach adapted from speculative execution. We show
that the standard PSO equations can be factored such that a set of speculative
positions can be found which will always include the position computed in the
next iteration.  By computing the value of the objective function for each of
the speculative positions at the same time the algorithm evaluates the
objective function for the current position, it is possible to know the
objective function values for both the current and the next iteration at the
same time.  We demonstrate this principle by implementation and show that it
produces exactly the same results as standard PSO, but two iterations at a
time.  The resulting implementation runs efficiently on large clusters where
the number of processors is much larger than a typical or reasonable number of
particles, producing better results in less wall-clock time.

@@

@@

Refactoring the PSO equations led us to find that speculative approaches are
possible in the parallelization of PSO.  SEPSO reproduces standard PSO exactly,
two iterations at a time, at the expense of requiring several times the number
of processors.  In this section we consider other speculative techniques
inspired by SEPSO that relax the requirement of exactly reproducing the
behavior of the original PSO algorithm.

@@

To make better use of the extra speculative evaluations, instead of choosing
the speculative child that matches the branch that the original PSO would have
taken, we take the child that has the best value.  The methodology is exactly
the same as with SEPSO except for the process of choosing which speculative
child to accept.  The only change needed in \alg{centralized} (see Appendix A)
is in step 7, where $\noeval{\s}_{t+1}$ with the best value is chosen from
$\noeval{\sset}_{t+1}$ instead of with the matching branch.  We call this
technique Pick Best.

@@

Because the SEPSO requires so many extra evaluations, a natural step to take is
to eliminate some of them.  If we could reliably predict which branch were
going to be taken, we could limit ourselves to one speculative evaluation per
particle instead of $2n+1$.  With a fixed number of processors, this would
allow us to greatly increase the swarm size relative to that needed in the
original speculative algorithm (e.g., with 120 processors, a standard
parallelization has a swarm of size 120, complete speculative evaluation has a
swarm of size 15, and pruning the evaluations to only one per particle allows a
swarm of size 60).  As not all of the branches are evaluated in any given
iteration, we call this technique pruning.

@@

In parallel PSO on long-running functions, the natural way to present results
is in terms of iterations, not function evaluations.  This is because when all
function evaluations at each iteration are performed concurrently, iterations
are the direct equivalent of wall-clock time.  Thus we report iterations in our
results instead of function evaluations.  But because SEPSO actually performs
two iterations of PSO at each ``iteration,'' and Social Promotion and Many
Iterations make the idea of ``iterations of PSO'' somewhat nebulus, we instead
call each ``iteration'' a ``time step.''

@@

When performing pruning in Pick Best Pruned and Social Promotion Pruned, there
are a large number of ways to prune speculative evaluations.  We experimented
with several, but present results for only one possible pruning.  The pruning
we present uses only the two branches where the $\nbest$ value was not updated:
\casepn\ and \casexn.  Those branches are convenient in that no messages are
needed from neighbors in order to produce the positions of the speculative
particles---in distributed frameworks using several rounds of communication
(see Appendix A), one of the rounds of communication can be dropped entirely.
Pruning all but these branches also allows the use of arbitrarily dense
topologies, as the number of speculative particles is no longer dependent on
the number of neighbors the particle has.

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{sphere20}
  \caption{Function Sphere with 20 dimensions, comparing a pertinent subset of
  possible methods.  Each method performs one evaluation on each of 240
  processors per time step.}
  \label{fig:sphere-20}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{rastrigin}
  \caption{Function Rastrigin with 20 dimensions.  Each method performs one
  evaluation on each of 240 processors per time step.}
  \label{fig:rastrigin}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank6}
  \caption{Function Griewank with 20 dimensions.  Each method performs one
  evaluation on each of 800 processors per time step.  Instead of showing
  average function value, we show the percentage of runs that have found the
  global optimum by each iteration.  All algorithms use the Ring topology.}
  \label{fig:basic-griewank1}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank1}
  \caption{Function Griewank with 20 dimensions.  Each method performs one
  evaluation of the objective function on each of 240 processors per time step.
  All algorithms use the Ring topology.}
  \label{fig:basic-griewank2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank3}
  \caption{Function Griewank with 20 dimensions, comparing Pick Best with
  results from \fig{basic-griewank2}.  Each method performs one evaluation on
  each of 240 processors per time step.  All algorithms use the Ring topology.}
  \label{fig:griewank-pickbest}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank4}
  \caption{Function Griewank with 20 dimensions, comparing pruning with the
  best results from \fig{griewank-pickbest}.  Each method performs one
  evaluation on each of 240 processors per time step.  All algorithms use the
  Ring topology.}
  \label{fig:griewank-pruned}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank5}
  \caption{Function Griewank with 20 dimensions, comparing speculating several
  iterations ahead with selections from \fig{griewank-pruned}.  Each method
  performs one evaluation on each of 240 processors per time step.  All
  algorithms use the Ring topology.}
  \label{fig:griewank-manyiters}
\end{figure}

@@

Bohachevsky is a unimodal function best optimized with a Complete swarm.  It is
similar to Griewank in that there is a global optimum with a value of 0, and
the swarm sometimes finds it and sometimes does not.  Thus we present a graph
similar to those of Griewank, because ``average value'' graphs have a
misleading flat line.  We used 480 processors to optimize this function.  In
\fig{bohachevsky} we show a plot with a few pertinent methods, while
\tabref{bohachevsky-20} shows all of the results.  All of our speculative
approaches found the optimum much quicker than Standard with a Random topology.
However, SEPSO was slower than Standard Complete and got stuck 25\% of the
time.  Pick Best, Pick Best Pruned, and Many Iterations all outperformed
Standard Complete, with Many Iterations finding the optimum about twice as
fast.

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{bohachevsky}
  \caption{Function Bohachevsky with 20 dimensions.  Each method performs one
  evaluation on each of 480 processors per time step.}
  \label{fig:bohachevsky}
\end{figure}

@@

\begin{table}
  \caption{Average speed up, comparing the number of time steps to completion
  for the best speculative topology to the time steps to completion for the
  best standard topology.  All functions have 50 dimensions.}
  \label{tab:50dims-summary}
  \centering
  \begin{tabular}{|l|r|}
	\hline
	Function&Speed Up Factor\\
	\hline
	\hline
	Rastrigin&N/A\\
	\hline
	Sphere&1.84\\
	\hline
	Schwefel&2.51\\
	\hline
	Griewank&6.08\\
	\hline
	Bohachevsky&3.07\\
	\hline
  \end{tabular}
\end{table}

@@

At 500 dimensions the performance of constricted PSO on benchmark functions
becomes rather dismal.  In almost all cases, Standard fails to make any
significant progress.  Our search for an explanation seems to say that the
space is too large for particles to converge to the same point, so all updates
to the best position found come from a single particle wandering in the space.
Every time that single particle finds a new best position, its velocity
contracts, so the particle is unable to make significant progress on its own.
In constricted PSO it is necessary to have a collection of particles exploring
a promising location to keep the particles' velocities from contracting
prematurely, and in 500 dimensions the space is too large to get the collection
of particles to the same location.

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{sphere500}
  \caption{Function Sphere with 500 dimensions.  Each method performs one
  evaluation on each of 800 processors per time step.}
  \label{fig:sphere500}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank500}
  \caption{Function Griewank with 500 dimensions.  Each method performs one
  evaluation on each of 800 processors per time step.}
  \label{fig:griewank500}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{rastrigin500}
  \caption{Function Rastrigin with 500 dimensions.  Each method performs one
  evaluation on each of 800 processors per time step.}
  \label{fig:rastrigin500}
\end{figure}

@@

\section*{Appendix A---Implementing Speculative Evaluation}
\label{sec:implementation}

It is not trivial in some parallel architectures to determine which speculative
position was the correct next position of each particle.  In this section we
discuss in detail some important considerations in the implementation of our
methods.  First we discuss the relatively easy case of a centralized parallel
PSO algorithm with a master computer and many slaves.  In such an architecture,
the master keeps track of all necessary information with only trivial message
passing needed.  Then we discuss the more complicated case of a distributed
algorithm, where each particle is on its own and needs to send and receive
messages to and from other particles.  Finally we discuss the further
complications of a dynamic topology such as Random, where a particle's
neighbors change from one iteration to another.

In distributed frameworks, synchronizing all of the machines for a round of
message passing can be expensive.  The method just described uses three rounds
of message passing for every two iterations (corresponding to steps 2, 5 and 10
in \alg{centralized}).  It is possible to perform speculative evaluation in PSO
with only one round of communication per two iterations.  However, the
methodology is tedious and distracting from the present discussion, so we defer
its description to Appendix B.

\section*{Appendix B---Alternate form of message passing}
